---
title: Making Sense of $1.3T in Student Loan Debt - An Analysis of Student Loan Hero
  Users and Predictive Default Model
author: "Trevor Ford"
date: "April 2, 2017"
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
  html_document:
    toc: yes
    toc_depth: '2'
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
library(dplyr) #dplyr package import
library(lubridate) #lubridate (For working with date variables)
library(ggplot2) # for plotting
library(plyr)
library(magrittr) 
library(RecordLinkage)
library(stringdist)
library(readr)
library(choroplethr)
library(choroplethrMaps)
setwd('C:\\Users\\Trevor\\Documents\\slhloans')
slhloans <- read.csv("loans_final_userid.csv")
```

# Introduction

Student loan debt is a $1.4T problem facing over 44 million Americans.  The average 2016 graduate left college with over $37,000 of debt and is burdened with over $350 a month in student loan payments.  Student Loan Hero was conceived as a resource to help student loan borrowers understand the various student loan repayment schemes, navigate the numerous federal programs designed to help borrowers with their loans, and recommend ways for borrowers to save money on their loans.  Student Loan Hero developed a proprietary tool that aggregates all of a borrower's loans and displays the details all in one place, then generates a customized repayment plan based on each borrower's unique circumstances.  To date, the tool has analyzed over $2 billion worth of student loan debt and helped 50,000 plus borrowers best pay off their student loans.  This course offered the perfect opportunity to explore the vast amount of borrower data gathered in the three years Student Loan Hero has been operating.

## The Goal

Student loan default rates are on the rise [elaborate here, why default is a bad outcome for borrowers, etc]... By using a binary classification machine learning algorithm we hope to be able to predict whether a given borrower is likely to default on their student loans.  The business application of this prediction is if a borrower is likely to default on their student loans, Student Loan Hero has the opportunity to connect with high-risk borrowers and help borrowers take appropriate action to avoid default.


# Datasets
##Student Loan Hero User Dataset

The private Student Loan Hero User dataset contains data from 24,000 anonymized student loans. 

```{r slhloans}
names(slhloans)
```


## Carnegie University Classifications

The public Carnegie University Classifications dataset is a framework for recognizing and describing institutional diversity in U.S. higher education.  The dataset contains detailed data about all Title IV colleges and universities.

# Preparing and Cleaning Datasets

Since the goal of this project is to predict the likelihood a borrower will default on their loans by use of a binary classification machine learning model it was imperative to narrow the number of levels in a number of factors in the Student Loan Hero dataset.

## Status
The most important factor to address is the loan "status", a factor containing a number of statuses a borrower's loan could be in.
```{r}
str(slhloans$Status)
head(levels(factor(slhloans$Status)))
```
This factor with 149 levels needed to be reduced to just two levels for the purposes of the desired machine learning algorithm.  A borrower's loan could have one of two statuses, either in "Repayment" or in "Default".  This meant first breaking down the statuses by frequency, identifying duplicate but misnamed levels, and merging levels that had similar meanings.

```{r, echo=FALSE}
slhloans$Status <- as.character(slhloans$Status)

slhloans$Status[grep(".*defer.*", slhloans$Status,ignore.case = TRUE)] <- "RPM"
slhloans$Status[grep(".*Originated.*|.*in school.*", slhloans$Status,ignore.case = TRUE)] <- "RPM"
slhloans$Status[grep(".*repayment.*", slhloans$Status,ignore.case = TRUE)] <- "RPM"
slhloans$Status[grep(".*grace.*", slhloans$Status,ignore.case = TRUE)] <- "RPM"
slhloans$Status[grep(".*default.*", slhloans$Status,ignore.case = TRUE)] <- "Default"
slhloans$Status[grep(".*forbearance.*|.*forb.*", slhloans$Status,ignore.case = TRUE)] <- "RPM"
slhloans$Status[grep(".*bankruptcy.*", slhloans$Status,ignore.case = TRUE)] <- "Default"
slhloans$Status[grep(".*delinq.*", slhloans$Status,ignore.case = TRUE)] <- "Default"
slhloans$Status[grep(".*CANCEL.*|.*Custom.*|.*interest only.*", slhloans$Status,ignore.case = TRUE)] <- "RPM"

```

```{r}
levels(factor(slhloans$Status))
```
## Major

Cleaning the major's obtained by borrowers was necessay to undertake since the non-standardized factor had over 1,300 different entries, the majority of which were duplicative in nature (for example, one borrower's degree read "International Studies" while another's read "Int'l Studies").
```{r}
str(slhloans$Major)
```

Condensing the majors to the most frequently occuring and a basket titled "others" for any other majors made this factor much easier to work with.

```{r}
levels(slhloans$Major) <- gsub(".*Law.*|.*Lawyer.*|.*attorney.*|.*JD.*|.*Juris Doctor.*|.*Paralegal.*", "Law", levels(slhloans$Major))
levels(slhloans$Major) <- gsub(".*Business.*|.*Financ.*|.*Account.*|.*Economic.*|.*Marketing.*|.*Human Resources.*", "Business", levels(slhloans$Major))
levels(slhloans$Major) <- gsub(".*Philosophy.*|.*Social.*|.*Politic.*|.*English.*|.*History.*|.*Sociology.*|.*International.*|.*General Studies.*|.*Fashion.*", "Liberal Arts", levels(slhloans$Major))
levels(slhloans$Major) <- gsub(".*Engineer.*", "Engineering", levels(slhloans$Major))
levels(slhloans$Major) <- gsub(".*Medicine.*|.*MD.*|.*Physician*|.*Doctor.*|.*Pharmacy.*|.*Medical.*|.*medicine.*", "Higher Medical Degree", levels(slhloans$Major))
levels(slhloans$Major) <- gsub(".*Teach.*|.*Education.*|.*Teach.*", "Education", levels(slhloans$Major))
levels(slhloans$Major) <- gsub(".*Technology.*|.*Computer.*|.*Systems.*", "Computer Science", levels(slhloans$Major))
levels(slhloans$Major) <- gsub(".*Communication.*|.*Relations.*", "Communications", levels(slhloans$Major))
levels(slhloans$Major) <- gsub(".*Science.*|.*Biology.*|.*Chemistry.*|.*Geology.*|.*Bioengineer.*|.*Physics.*|.*Mathematics.*", "Sciences", levels(slhloans$Major))
levels(slhloans$Major) <- gsub(".*Psychology.*", "Psychology", levels(slhloans$Major))
levels(slhloans$Major) <- gsub("Art|.*Graphic.*|.*Film.*|.*Drama.*|.*Music.*|.*Photography.*", "Arts", levels(slhloans$Major))
levels(slhloans$Major) <- gsub(".*Nurs.*|.*nurs.*", "Nursing", levels(slhloans$Major))

OtherMajors <- !(slhloans$Major %in% c("Business","Sciences", "Higher Medical Degree", "Engineering", "Liberal Arts", "Law", "Education", "Nursing", "Psychology", "Communications", "MBA"))

slhloans$Major[OtherMajors]<- "Other"

slhloans$Major <- factor(slhloans$Major)

levels(slhloans$Major)
```


## Profession
Cleaning the profession's of borrowers again was important since to start borrowers had self-identified into over 1800 different professions.  
```{r}
str(slhloans$Profession)
```
By first merging duplicative or identical professions (example: accountant, certified public accountant) and then analyzing the most frequently occuring professions the profession level was narrowed to ten of the top professions and a catchall category of "Other" for all other professions.
```{r}
levels(slhloans$Profession) <- gsub(".*Lawyer.*|.*attorney.*", "Attorney", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Engineer.*", "Engineer", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Teacher.*", "Teacher", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Nurse.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Doctor.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Therapist.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Physician.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Dentist.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Dental.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Physician's.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Psychologist.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Psychiatrist.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Medical.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Chiropractor.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Pharmacist.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Nursing.*", "Doctor/Nurse/Pharmacist", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Computer.*", "Computer/Tech", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Developer.*", "Computer/Tech", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*IT.*", "Computer/Tech", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Teacher.*", "Teacher", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Librarian.*", "Teacher", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Professor.*", "Teacher", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Accounting.*", "Accountant", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Student.*|.*Resident.*|.*Researcher.*|.*student.*", "Student", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Designer.*|.*Graphic.*", "Designer", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Business.*|.*Manager.*|.*Marketing.*|.*Financial.*|.*Analyst.*|.*Insurance.*|.*Consultant*.|.*Resources.*|.*Consulting.*", "General Business", levels(slhloans$Profession))
levels(slhloans$Profession) <- gsub(".*Retail.*|.*Cashier.*", "Retail", levels(slhloans$Profession))

OtherProfessions <- !(slhloans$Profession %in% c("Doctor/Nurse/Pharmacist","General Business", "Engineer", "Attorney", "Teacher", "Computer/Tech", "Accountant", "Student", "Designer", "Retail"))

slhloans$Profession[OtherProfessions]<- "Other"

slhloans$Profession <- factor(slhloans$Profession)

```

```{r}
levels(factor(slhloans$Profession))
```

## Other Factors

By using domain knowledge the "Servicer" factor was removed as it has no bearing on whether a borrower will default or not.

#Exploratory Data Analysis

In examning the new cleaned data set, we can see we have a fairly unbalanced data set with the majority of borrowers in "repayment" and only a handful in "default".
```{r}
slhloans%>%group_by(Status)%>% dplyr::summarise(Count=n())

```

[1.4% calculation]

##Loans in repayment and default by degree type
```{r}
ggplot(slhloans, aes(Education.Degree, fill=Status))+geom_bar()+theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

In order to better understand where users are in their journey of paying off their student loan debt, a column calculating the difference between the original loan's balance and the current loan's balance was added.

```{r}
#adding a new calculated column = difference between original and current principal

slhloans <- mutate(slhloans, Difference=Current.Principal-Original.Principal)

summary(slhloans$Difference)
```

To ensure loans where there is currently a balance present are used, the previously created column "difference" 

```{r}
#Using only records where difference is positive, and rate is >0
slhloans <- filter(slhloans, Difference>=0&Rate>0)
```

To ensure only loans where there was an initial balance added by the user, any loans with an original balance of 0 were removed.

```{r}

#getting rid of records where original principal is 0
slhloans <- filter(slhloans, Original.Principal>0)
```



```{r}
ggplot(slhloans, aes(Education.Degree, fill=Status))+geom_bar()+theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Since the database where the loans was held had a non-standardized date field, using lubridate the dates were converted to a format where calculations could be performed on them.

```{r}
#Lubridating the date field from character to true date field
slhloans$Loan.Disbursement.Date <- mdy(slhloans$Loan.Disbursement.Date)

```

##Loan disbursement distribution by year

Taking a look at the distribution of years when the users' loans were disbursed 
```{r echo=FALSE}
#yearwise breakdown of loans disbursed
yearwise <- slhloans%>%group_by(year(Loan.Disbursement.Date))%>%dplyr::summarize(Count=n())
```

```{r}
yearwise_desc <- arrange(yearwise, desc(Count))

colnames(yearwise) <- c("year", "count")

ggplot(yearwise, (aes(x = year, y = count))) + geom_bar(stat="identity")
```


##Student Loan Hero users by state
```{r echo=FALSE}
slhloans_states <- slhloans%>%group_by(State.of.Residency)%>%dplyr::summarise(Status=n())

summary(slhloans_states)
```


```{r}
ggplot(slhloans_states, aes(slhloans_states$State.of.Residency)) + geom_bar() + facet_grid(slhloans_states$State.of.Residency~ .)

```


```{r}
#Adding new column for each user's original balance sum (since users can have multiple loans in their name)
slhloans$Original.Principal[is.na(slhloans$Original.Principal)] <- 0
slhloans$Current.Principal[is.na(slhloans$Current.Principal)] <- 0

slhloansv2 <- slhloans %>% group_by(User.ID..) %>% mutate(origtotalbalance=sum(Original.Principal))
slhloansv2 <- slhloans %>% group_by(User.ID..) %>% mutate(currenttotalbalance=sum(Current.Principal))%>% mutate(origtotalbalance=sum(Original.Principal))

slhloansv2$progress <- mutate(slhloansv2, progress = currenttotalbalance / origtotalbalance)

```

##Carnegie Mellon University Classifications
The Carnegie Mellon University Classification dataset is a framework for categorizing and classifying United States institues of higher education.  All Title IV colleges and universities are listed along with accompanying classifying data including attributes such as enrollment profiles (demographics, student statuses, etc), university size and setting, any unique university characteristics such as whether the school is a technical or vocational school, historically black colleges and universities, women's only universities, etc.
```{r}
cleanloans <- read.csv("slhloansclean.csv") #this is the post-cleaned version of "loans"
```

Once the Carnegie Mellon University Classification dataset was imported, the classifiers that were to be used for further analysis and matched up with the universities of SLH users were imported, including data such as the location (state and region) of university, type of university (is this a medical school, liberal arts school, womens school, etc).

```{r}
univ <- read.csv("collegedetails.csv")
cleanloans$univctrl <- univ[match(cleanloans$College, univ$NAME),4]
cleanloans$univstate <- univ[match(cleanloans$College, univ$NAME),3]
cleanloans$univobereg <- univ[match(cleanloans$College, univ$NAME),5]
cleanloans$univlocale <- univ[match(cleanloans$College, univ$NAME),6]
cleanloans$univenrprofile <- univ[match(cleanloans$College, univ$NAME),8]
cleanloans$univmedical <- univ[match(cleanloans$College, univ$NAME),9]
cleanloans$univhbcu <- univ[match(cleanloans$College, univ$NAME),10]
cleanloans$univtribal <- univ[match(cleanloans$College, univ$NAME),11]
cleanloans$univhsi <- univ[match(cleanloans$College, univ$NAME),12]
cleanloans$univwomens <- univ[match(cleanloans$College, univ$NAME),14]
cleanloans$univlibarts <- univ[match(cleanloans$College, univ$NAME),15]
```

##More Cleaning

```{r echo="false"}
levels(cleanloans$Major) <- gsub(".*Law.*|.*Lawyer.*|.*attorney.*|.*JD.*|.*Juris Doctor.*|.*Paralegal.*", "Law", levels(cleanloans$Major))
levels(cleanloans$Major) <- gsub(".*Business.*|.*Financ.*|.*Account.*|.*Economic.*|.*Marketing.*|.*Human Resources.*", "Business", levels(cleanloans$Major))
levels(cleanloans$Major) <- gsub(".*Philosophy.*|.*Social.*|.*Politic.*|.*English.*|.*History.*|.*Sociology.*|.*International.*|.*General Studies.*|.*Fashion.*", "Liberal Arts", levels(cleanloans$Major))
levels(cleanloans$Major) <- gsub(".*Engineer.*", "Engineering", levels(cleanloans$Major))
levels(cleanloans$Major) <- gsub(".*Medicine.*|.*MD.*|.*Physician*|.*Doctor.*|.*Pharmacy.*|.*Medical.*|.*medicine.*", "Higher Medical Degree", levels(cleanloans$Major))
levels(cleanloans$Major) <- gsub(".*Teach.*|.*Education.*|.*Teach.*", "Education", levels(cleanloans$Major))
levels(cleanloans$Major) <- gsub(".*Technology.*|.*Computer.*|.*Systems.*", "Computer Science", levels(cleanloans$Major))
levels(cleanloans$Major) <- gsub(".*Communication.*|.*Relations.*", "Communications", levels(cleanloans$Major))
levels(cleanloans$Major) <- gsub(".*Science.*|.*Biology.*|.*Chemistry.*|.*Geology.*|.*Bioengineer.*|.*Physics.*|.*Mathematics.*", "Sciences", levels(cleanloans$Major))
levels(cleanloans$Major) <- gsub(".*Psychology.*", "Psychology", levels(cleanloans$Major))
levels(cleanloans$Major) <- gsub("Art|.*Graphic.*|.*Film.*|.*Drama.*|.*Music.*|.*Photography.*", "Arts", levels(cleanloans$Major))
levels(cleanloans$Major) <- gsub(".*Nurs.*|.*nurs.*", "Nursing", levels(cleanloans$Major))

OtherMajors <- !(cleanloans$Major %in% c("Business","Sciences", "Higher Medical Degree", "Engineering", "Liberal Arts", "Law", "Education", "Nursing", "Psychology", "Communications", "MBA"))

cleanloans$Major[OtherMajors]<- "Other"

cleanloans$Major <- factor(cleanloans$Major)

levels(cleanloans$Major)
```
```{r echo="false"}

cleanloans$Status[grep("forbearance",cleanloans$Status,ignore.case = TRUE)] <- "FBR"

levels(factor(cleanloans$Status)) #prints number of unique levels in status column

#further reducing duplicative levels
cleanloans$Status[grep("deferment", cleanloans$Status,ignore.case = TRUE)] <- "DFR"
cleanloans$Status[grep("repayment", cleanloans$Status,ignore.case = TRUE)] <- "RPM"
cleanloans$Status[grep("grace", cleanloans$Status,ignore.case = TRUE)] <- "grace"
cleanloans$Status[grep("default", cleanloans$Status,ignore.case = TRUE)] <- "default"

levels(cleanloans$Status) <- gsub(".*DEFERRED.*", "DFR", ignore.case = TRUE, levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub(".*DEFERMENT.*", "DFR", ignore.case = TRUE, levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub(".*repayment.*", "RPM", ignore.case = TRUE, levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub(".*Grace.*", "DFR", ignore.case = TRUE, levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub(".*default.*", "Default", ignore.case = TRUE, levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub(".*forbearance.*", "DFR", ignore.case = TRUE, levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub(".*bankruptcy.*", "Default", ignore.case = TRUE, levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub(".*delinq.*", "Default", ignore.case = TRUE, levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub(".*rpm.*", "RPM", ignore.case = TRUE, levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub(".*defer.*", "DFR", ignore.case = TRUE, levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub(".*Forb.*", "DFR", ignore.case = TRUE, levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub(".*In school.*", "In School", ignore.case = TRUE, levels(cleanloans$Status))


#levels(cleanloans$Status) <- gsub("DEFERRED", "DFR", levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub("LOAN ORIGINATED", "IN SCHOOL", levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub("FBR", "DFR", levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub("grace", "DFR", levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub("DFR", "DFR/GRC", levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub("DFR/GRC/GRC", "DFR/GRC", levels(cleanloans$Status))

levels(factor(cleanloans$Status)) #checking number of levels now

```
```{r echo="FALSE"}
#Cleaning up the Professions
head(levels(cleanloans$Profession))

head(sort(table(cleanloans$Profession)))

levels(cleanloans$Profession) <- gsub(".*Lawyer.*|.*attorney.*", "Attorney", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Engineer.*", "Engineer", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Teacher.*", "Teacher", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Nurse.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Doctor.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Therapist.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Physician.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Dentist.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Dental.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Physician's.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Psychologist.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Psychiatrist.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Medical.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Chiropractor.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Pharmacist.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Nursing.*", "Doctor/Nurse/Pharmacist", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Computer.*", "Computer/Tech", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Developer.*", "Computer/Tech", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*IT.*", "Computer/Tech", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Teacher.*", "Teacher", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Librarian.*", "Teacher", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Professor.*", "Teacher", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Accounting.*", "Accountant", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Student.*|.*Resident.*|.*Researcher.*|.*student.*", "Student", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Designer.*|.*Graphic.*", "Designer", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Business.*|.*Manager.*|.*Marketing.*|.*Financial.*|.*Analyst.*|.*Insurance.*|.*Consultant*.|.*Resources.*|.*Consulting.*", "General Business", levels(cleanloans$Profession))
levels(cleanloans$Profession) <- gsub(".*Retail.*|.*Cashier.*", "Retail", levels(cleanloans$Profession))
OtherProfessions <- !(cleanloans$Profession %in% c("Doctor/Nurse/Pharmacist","General Business", "Engineer", "Attorney", "Teacher", "Computer/Tech", "Accountant", "Student", "Designer", "Retail"))

cleanloans$Profession[OtherProfessions]<- "Other"

cleanloans$Profession <- factor(cleanloans$Profession)

levels(factor(cleanloans$Profession))

#removing servicer column - unnecessary for predicting the chance a borrower will default
cleanloans$Servicer <- NULL
```

##Segmenting Data

```{r}
#adding a new calculated column = difference between original and current principal
cleanloans <- mutate(cleanloans, Difference=Current.Principal-Original.Principal)

summary(cleanloans$Difference)

#Pulling records where difference is positive, and rate is >0
cleanloans <- filter(cleanloans, Difference>=0&Rate>0)

#str(cleanloans)

#deleting variables that are null or useless
#cleanloans <- select(cleanloans, -(Created.At:Loan.ID..),-X)

#getting rid of records where original principal is 0
cleanloans <- filter(cleanloans, Original.Principal>0)

#Lubridating the date field from character to true date field
cleanloans$Loan.Disbursement.Date <- mdy(cleanloans$Loan.Disbursement.Date)

#Adding new column for each user's original balance sum
cleanloans$Original.Principal[is.na(cleanloans$Original.Principal)] <- 0
cleanloans <- cleanloans %>% group_by(User.ID..) %>% mutate(origtotalbalance=sum(Original.Principal))
#Adding new column for each user's current balance sum
cleanloans$Current.Principal[is.na(cleanloans$Current.Principal)] <- 0
cleanloans <- cleanloans %>% group_by(User.ID..) %>% mutate(currenttotalbalance=sum(Current.Principal))%>% mutate(origtotalbalance=sum(Original.Principal))


#NEW R CHUNK?
#setup column for is current principal > original principal

#change borrower home state to obereg to match the university state obereg

levels(cleanloans$State.of.Residency) <- gsub("VT|CT|ME|MA|NH|RI", "1", levels(cleanloans$State.of.Residency))
levels(cleanloans$State.of.Residency) <- gsub("DE|DC|MD|NJ|NY|PA", "2", levels(cleanloans$State.of.Residency))
levels(cleanloans$State.of.Residency) <- gsub("IL|IN|MI|OH|WI", "3", levels(cleanloans$State.of.Residency))
levels(cleanloans$State.of.Residency) <- gsub("IA|KS|MN|MO|NE|ND|SD", "4", levels(cleanloans$State.of.Residency))
levels(cleanloans$State.of.Residency) <- gsub("AL|AR|FL|GA|KY|LA|MS|NC|SC|TN|VA|WV", "5", levels(cleanloans$State.of.Residency))
levels(cleanloans$State.of.Residency) <- gsub("AZ|NM|OK|TX", "6", levels(cleanloans$State.of.Residency))
levels(cleanloans$State.of.Residency) <- gsub("CO|ID|MT|UT|WY", "7", levels(cleanloans$State.of.Residency))
levels(cleanloans$State.of.Residency) <- gsub("AK|CA|HI|NV|OR|WA", "8", levels(cleanloans$State.of.Residency))
levels(cleanloans$State.of.Residency) <- gsub("AS|FM|GU|MH|MP|PR|PW|VI", "9", levels(cleanloans$State.of.Residency))

#Cleaning up Loan Names to just 2 type: Federal Loans or Private Loans
head(levels(cleanloans$Name))

levels(cleanloans$Name) <- gsub(".*Private.*", "Private", levels(cleanloans$Name))
levels(cleanloans$Name) <- gsub(".*Fargo.*", "Private", levels(cleanloans$Name))
levels(cleanloans$Name) <- gsub(".*Sallie.*", "Private", levels(cleanloans$Name))
levels(cleanloans$Name) <- gsub(".*Stafford.*|.*STAFFORD.*|.*stafford.*", "Federal", levels(cleanloans$Name))
levels(cleanloans$Name) <- gsub(".*Direct.*|.*direct.*|.*DIRECT.*", "Federal", levels(cleanloans$Name))
levels(cleanloans$Name) <- gsub(".*Federal.*|.*federal.*|.*FEDERAL.*", "Federal", levels(cleanloans$Name))
levels(cleanloans$Name) <- gsub(".*Signature.*|.*signature.*|.*SIGNATURE.*", "Federal", levels(cleanloans$Name))
levels(cleanloans$Name) <- gsub(".*subsidized.*|.*Subsidized.*|.*SUBSIDIZED.*", "Federal", levels(cleanloans$Name))
levels(cleanloans$Name) <- gsub(".*PLUS.*|.*plus.*|.*Plus.*", "Federal", levels(cleanloans$Name))
levels(cleanloans$Name) <- gsub(".*FFEL.*|.*ffel.*|.*Ffel.*", "Federal", levels(cleanloans$Name))
levels(cleanloans$Name) <- gsub(".*CONSOLIDATION.*|.*consolidation.*|.*Consolidation.*", "Federal", levels(cleanloans$Name))

OtherLoans <- !(cleanloans$Name %in% c("Private","Federal"))

cleanloans$Name[OtherLoans]<- "Private"

cleanloans$Name <- factor(cleanloans$Name)

levels(factor(cleanloans$Name))

sort(table(cleanloans$Name))

#Reducing Loan Statuses to just two
table(cleanloans$Status)
levels(cleanloans$Status) <- gsub("DFR/GRC", "Repayment", levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub("CANCELLED", "Repayment", levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub("In School", "Repayment", levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub("Interest Only", "Repayment", levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub("IN SCHOOL", "Repayment", levels(cleanloans$Status))
levels(cleanloans$Status) <- gsub("RPM", "Repayment", levels(cleanloans$Status))
```
##Next Data Points
Now that we've 
```{r}
cleanloans$User.DOB <- mdy(cleanloans$User.DOB)
cleanloans$User.ID.. <- as.factor(cleanloans$User.ID..)
cleanloans$Loan.Disbursement.Date <- mdy(cleanloans$Loan.Disbursement.Date)
cleanloans$Joint.Federal.Income.Tax. <- as.character(cleanloans$Joint.Federal.Income.Tax.)

#In order to use choropleth maps, I need the state names
data("state.regions")

#Loans per state
state_loans <- cleanloans%>%group_by(abb=univstate)%>%dplyr::summarize(value=n())
state_loans <- state_loans[complete.cases(state_loans),]
state_loans <- left_join(state_loans, state.regions, by='abb')
state_choropleth(state_loans, title="Loans Per State")

#Defaulted Loans as per status
status <- 'Default'
state_loans <- cleanloans%>%filter(Status==status)%>%group_by(abb=univstate)%>%dplyr::summarize(value=n())
state_loans <- state_loans[complete.cases(state_loans),]
state_loans <- left_join(state_loans, state.regions, by='abb')
state_choropleth(state_loans,title=paste0("Loans As Per Status:",status))

#Loans as per rate of interest
state_loans <- cleanloans%>%group_by(abb=univstate)%>%dplyr::summarize(value=mean(Rate))
state_loans <- state_loans[complete.cases(state_loans),]
state_loans <- left_join(state_loans, state.regions, by='abb')
state_choropleth(state_loans,title="Average Rate of Interest")
```
## TEXT HERE
```{r}
#AVG INTEREST RATES by Fixed vs Variable Interest Rates
#Type <- 'Fixed'
loan_rates <- cleanloans%>%group_by(Type)%>%dplyr::summarize(value=mean(Rate))

ggplot(loan_rates, (aes(x = Type, y = value))) + geom_bar(stat="identity")

#AVG INTEREST RATES by Status
loan_rates_status <- cleanloans%>%group_by(Status)%>%dplyr::summarize(value=mean(Rate))

ggplot(loan_rates_status, (aes(x = Status, y = value))) + geom_bar(stat="identity")

#AVG INTEREST RATES by Loan Type (Federal or Private)
loan_rates_name <- cleanloans%>%group_by(Name)%>%dplyr::summarize(value=mean(Rate))

ggplot(loan_rates_name, (aes(x = Name, y = value))) + geom_bar(stat="identity")

#AVG BALANCE by Loan Type (Federal or Private)
loan_balance_name <- cleanloans%>%group_by(Name)%>%dplyr::summarize(value=mean(Original.Principal))

ggplot(loan_balance_name, (aes(x = Name, y = value))) + geom_bar(stat="identity")

#AVG BALANCE by Rate Type (Fixed vs Variable)
loan_balance_type <- cleanloans%>%group_by(Type)%>%dplyr::summarize(value=mean(Original.Principal))

ggplot(loan_balance_type, (aes(x = Type, y = value))) + geom_bar(stat="identity")

#AVG BALANCE by Status
loan_balance_status <- cleanloans%>%group_by(Status)%>%dplyr::summarize(value=mean(Original.Principal))

ggplot(loan_balance_status, (aes(x = Status, y = value))) + geom_bar(stat="identity")

#Status by Majors
ggplot(cleanloans, aes(Major, fill=Status))+geom_bar()+theme(axis.text.x = element_text(angle = 90, hjust = 1))

#Fixed/Variable Loans by Major
ggplot(cleanloans, aes(Major, fill=Type))+geom_bar()+theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

#Predicting Default Rates
##Machine Learning Algorithm - Random Forest
The goal of our default prediction model was to determine whether a given borrower would fall into one of two statuses, the "Default" status or the "Repayment" tatus.  In determining the best machine learning algorithm to apply to our classifier model Random Forest was chosen for its   

##Variable Importance


##Variable Importance
```{r echo="FALSE"}

```

#Acknowledgements
I'd like to thank Dhiraj Khanna for his sustained and continued support in preparing this project and report and for the numerous suggestions surrounding how to work with unbalanced data sets. Invaluable knowledge in how to approach, work in R, random forest models..



